Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study

### 2025-06-19 inst/paper/main.tex
Add discussion points

\item Digital tracing has capacity for large-scale contact tracing (pingdemic), also rapid tests are a complementary approach for limiting spread.

@@

\item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic.
\item In this study we have parameterised the transmission model with parameter from avian influenza subtypes (H5N1 and H7N9) and a influenza subtype that caused an epidemic in the recent past (H1N1), and response delays from SARS and COVID-19 outbreaks. However, these results can be used to inform other influenza subtypes that exhibit similar epidemiological characteristics in transmissibility and severity from past outbreaks (H2N2 and H3N2) and still exist in animal reservoirs for potential spillovers back into human populations (ref).

@@

\item The reproduction number for H5N1 (including clade 2.3.4.4b) and H7N9 influenza subtypes have been estimated to be far below 1 (Ward et al., Garg et al., 2025). Therefore, the transmissibility scenarios explored in this study are speculative and assume that pathogen evolution increases human-to-human transmission potential. If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions. If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence isolation and contact tracing while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
\item This study has evaluated the ability to control the outbreak of an infectious disease independent on its severity. Infection severity and fatality risks are of course critical when formulating a pandemic response plan. By understanding the efficacy of isolation and contact trace for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.

>>>>

\item Digital tracing has capacity for large-scale contact tracing, also rapid tests are a complementary approach for limiting spread.

@@

\item Mention strengths: interpretability, focus on average impact, speed
\item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic -- possibly beyond scope if focused on burden.
\item In this study we have parameterised the transmission model with parameter from avian influenza subtypes (H5N1 and H7N9) and an influenza subtype that caused an epidemic in the recent past (H1N1), and response delays from SARS and COVID-19 outbreaks. However, these results can be used to inform other influenza subtypes that exhibit similar epidemiological characteristics in transmissibility and severity from past outbreaks (H2N2 and H3N2) and still exist in animal reservoirs for potential spillovers back into human populations (ref).

@@

\item The reproduction number for H5N1 (including clade 2.3.4.4b) and H7N9 influenza subtypes have been estimated to be far below 1 (Ward et al., Garg et al., 2025). Therefore, the transmissibility scenarios explored in this study are mostly focused on planning purposes. If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions. If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence isolation and contact tracing while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
\item This study has evaluated the ability to control the outbreak of an infectious disease independent on its severity. Infection severity and fatality risks are of course critical when formulating a pandemic response plan. By understanding the efficacy of isolation and contact trace for a given epidemiological characteristic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.
